Model-order reduction of non-linear time-dependent neutronics via POD-Galerkin projection and DEIM

### 2022-07-10 paper_/main.tex
addressed reviewer comments

%------------------------------------------------------------------------------
% 
%------------------------------------------------------------------------------
\begin{abstract}
Presented here is an application of model-order reduction to non-linear neutronic transients.
The effort described extends previous work\cite{elzohery2021exploring} aimed to develop and assess a ROM framework that can be used for different purposes, such as design optimization and propagating input parameter uncertainties like nuclear data.
Here, the target applications is non-linear.
The ROM was built with the use of the POD-Galerkin projection method and the matrix version of the Discrete Empirical Interpolation Method (MDEIM) for improved efficiency.
The MDEIM is used to approximate the projection of the problem operator that otherwise must be computed at each non-linear iteration within each time step, which increases the ROM cost.
By applying the method to the LRA benchmark, speed ups from 6 to 40 were achieved depending on the full-order model (FOM) spatial fidelity.
The maximum relative error in the assemblies power was in the order of $1\times10^{-5}$\%.
Moreover, the ROM was parameterized by POD-greedy sampling, and a case study was performed where the kinetic data, i.e, the precursor decay constants and delayed-neutron fraction were assumed to be the model parameters of interest.
For the parameterized ROM, the maximum relative error in the assembly power was $1\times 10^{-3}$\%, 
The computational gain from the MDEIM-ROM was compared with that of a direct projection ROM and showed that employing the MDEIM is computationally advantageous, and the advantage grows with dimension of the problem.
The method was implemented in the open-source deterministic transport code Detran\cite{roberts2014advanced} where only access to the system operator was needed to construct the ROM, and; hence, the method is trivially invasive and could easily be implemented in many modern deterministic diffusion codes.

@@

In the application of our interest here, neutronic transients, the mathematical models are based on the diffusion approximation and the Boltzmann transport equation.
For realistic modeling, these neutron transport models need to be coupled with different physics models to account for their feedback.
For example, thermal-hydraulics feedback plays an essential role in reactor operation.
Changes in the fuel temperature alter the cross sections due to Doppler broadening \cite{stacey2018nuclear}, which in turn impacts the neutron population and hence the fission rate and the rate of heat generation.
Multi-physics modeling accounts for these different feedback mechanisms; however, it poses a challenge in terms of the computational cost,
since it entails solving coupled systems for different fields such as neutron flux, temperature, and flow rate.
Moreover, the neutron transport (or diffusion) equation is no longer a linear one due to the indirect dependence of the cross section on the neutron flux.
several approaches are used to solve such systems.
The most commonly used one is a serial coupling of two or more separate codes in such a way that the output of one at a given time step, such as the flux distribution, is used as an input to the other, e.g., the temperature (which can be used then to update the cross section and so on).
This procedure is commonly referred to as operator splitting or explicit coupling.
One drawback of operator splitting is that time integration requires small steps to avoid stability issues, thus leading to high computational time. 
Alternatively, tight coupling in time between the physics can be maintained by using an implicit scheme that solves a non-linear iteration in each time step, also increasing the computational time.
A modern algorithm such as Jacobian free Newton-Krylov (JFNK) \cite{gaston2009parallel} can be used to perform this implicit coupling.
Another popular approach is to use fixed-point iteration, with an implicit time scheme to iterate between difference physics within a time step until convergence is satisfied, which is computationally taxing.
The aim of this work is to use a reduced-order model to overcome the high computational expense of these non-linear models.
In particular, the ROM was developed using the POD-Galerkin projection.
The method belongs to the projection-based family and can be described as a physics-driven method since it requires knowledge of the original governing equation as will be shown in Sec.\ref{sec:POD}.
In a previous work \cite{elzohery2021exploring}, we applied the method to the time-dependent diffusion equation where there was no coupling with other models and hence the application was linear.
In this study, the developed ROM was extended to treat the non-linearity induced by multi-physics modeling.
To improve the computational efficiency of the developed ROM, a hyper-reduction technique known as the Discrete Empirical Interpolation Method (DEIM) was used in conjunction with the POD-Galerkin projection.
The primary purpose of using the DEIM is to obtain an offline-online decomposition of the ROM to avoid repetitive projection of the problem operator at each iteration within the time step.
The offline stage occurs only once and comprises the expensive step of the ROM.
On the other hand, the online stage involves solving a reduced number of equations at each time step with a computational cost independent of the FOM dimension and hence cheaper.
The LRA benchmark was used as a demonstrative application where the diffusion equation is coupled with a simplified thermal feedback model.

@@

%-----------------------------------------------------------------------------------------------------%
\section{The time-dependent diffusion equation}
\label{sec:fom}
In this work, the time dependent diffusion equation was used as the mathematical model of neutronic transients in nuclear reactors, i.e FOM, and for which an approximate ROM is sought.
The multi-group version of the equation is represented as
\begin{equation}
\begin{split}
\frac{1}{v_g} \frac{\partial\phi_g}{\partial t} &= 
\nabla. D_g\nabla\phi_g - \Sigma_{tg}\phi_g + \sum_{g^\prime=1}^{G}{\Sigma_{sg^\prime g}}\phi_{g^\prime} \\
&+ \chi_g(1-\beta)\sum_{g^\prime = 1}^{G}\nu\Sigma_{fg^\prime}\phi_{g^\prime}  
+ \chi_g\sum_{i=1}^{P} \lambda_i C_i \, ,
\end{split}
\label{time_diff}
\end{equation}
where 
\begin{equation*}
\frac{\partial C_i}{\partial t} = -\lambda_i C_i + \beta_i \sum_{g^\prime=1}^{G} \nu\Sigma_{fg^\prime} \phi_{g^\prime} \, ,
\end{equation*}
and $\phi_g$, $D_g$, $\Sigma_{ag}$, $\nu\Sigma_{fg}$ and $v_g$ are the neutron flux, diffusion coefficient, the absorption cross section, the production cross section and the average neutron velocity in group g, respectively. $\Sigma_{sg^\prime g}$ is the scattering cross section from group $g^\prime$ to group $g$ 
and $\chi_g$ is the probability that a fission neutron will be born in group $g$.
$C_i$ is the $i$th precursor concentration, $\beta_i$ is the fraction of delayed neutrons in group $i$, $\lambda_i$ is the $i$th group decay constant.
Upon spatial discretization, the resulting system of ODEs can be compactly represented in a matrix form as  
\begin{equation}
\frac{d\mathbf{y}(t)}{dt} = \mathbf{A}\mathbf{y}(t) \, ,
\label{eq:transient fom}
\end{equation}
where $\mathbf{y}\in\mathbb{R}^{N}$ is stacked vector of the flux and the precursors concentration, $\mathbf{A}\in\mathbb{R}^{N\times N}$ and $N = (I + G)\times n$ with $n$ is the number of spatial unknowns arising from discretization, $G$ is the number of energy groups, and $I$ is the number of precursors groups.
The operator $\mathbf{A}$ can be written in a block matrix form as
\begin{equation}
\mathbf{A} =  \left(\begin{array}{@{}c|c@{}}
\mathbf{L} & \mathbf{D} \\
\hline
\mathbf{F} & \mathbf{P}
\end{array}\right) \, .
\end{equation} 
where $\mathbf{L} \in \mathbb{R}^{(n\times G)\times (n\times G)}$ represents the diffusion operator,
$\mathbf{D} \in \mathbb{R}^{(n\times G)\times (n\times I)}$ is a block diagonal matrix representing the delayed neutrons production operator,
$\mathbf{P} \in \mathbb{R}^{(n\times I)\times ((n\times G)}$ is a diagonal matrix represents the precursors concentration loss operator, 
and $\mathbf{F}\in \mathbb{R}^{(n\times I)\times ((n\times I)}$ represents the precursors production operator.
%Thus, the system in Eq.~\ref{eq:transient fom} is the full-order model and it can be solved using a suitable time-scheme.
%However, it is expensive to solve such systems for fine spatial and time grids, since a linear system of a size proportional to the spatial discretization has to be solved for each time step.
%Hence, a reduced-order model to approximate it is justified.

@@

$\mathbf{\Phi}_1(t) = \mathbf{U}_1\mathbf{a}_1(t)$, $\mathbf{\Phi}(t)_2 = \mathbf{U}_2\mathbf{a}_2(t)$ and $\mathbf{C}(t) = \mathbf{U}_c\mathbf{a}_c(t)$, where $\mathbf{U}_1$, $\mathbf{U}_2$  and $\mathbf{U}_c$ are the POD bases of the fast group, thermal group and precursor concentration respectively, and $\mathbf{a}_1$, $\mathbf{a}_2$ and $\mathbf{a}_c$ are their respective temporal coefficients. 

>>>>

%------------------------------------------------------------------------------
% 
%------------------------------------------------------------------------------
\begin{abstract}
Presented here is an application of model-order reduction to non-linear neutronic transients.
The effort described extends previous work\cite{elzohery2021exploring} aimed to develop and assess a Reduced-order model(ROM) framework that can be used for different purposes, such as design optimization and propagating input parameter uncertainties like nuclear data.
Here, the target applications is non-linear.
The ROM was built with the use of the POD-Galerkin projection method and the matrix version of the Discrete Empirical Interpolation Method (MDEIM) for improved efficiency.
Where applications of the ROM to the diffusion equation have been introduced before \cite{}, this work focuses on treating the non-linearity induced by coupling with different physics.
Previous work has used the DEIM to approximate non-linear fielsa such as the temperature \cite{german2019application}.
Here, the MDEIM is used to approximate the projection of the problem operator that otherwise must be computed at each non-linear iteration within each time step, which increases the ROM cost.
By applying the method to the LRA benchmark, speed ups from 6 to 40 were achieved depending on the full-order model (FOM) spatial fidelity.
The maximum relative error in the assemblies power was in the order of $1\times10^{-5}$\%.
Moreover, the ROM was parameterized by POD-greedy sampling, and a case study was performed where the kinetic data, i.e, the precursor decay constants and delayed-neutron fraction were assumed to be the model parameters of interest.
For the parameterized ROM, the maximum relative error in the assembly power was $1\times 10^{-3}$\%, 
The computational gain from the MDEIM-ROM was compared with that of a direct projection ROM and showed that employing the MDEIM is computationally advantageous, and the advantage grows with dimension of the problem.
The method was implemented in the open-source deterministic transport code Detran\cite{roberts2014advanced} where only access to the system operator was needed to construct the ROM, and; hence, the method is trivially invasive and could easily be implemented in many modern deterministic diffusion codes.

@@

In the application of our interest here, neutronic transients, the mathematical models are based on the diffusion approximation and the Boltzmann transport equation.
For realistic modeling, these neutron transport models need to be coupled with different physics models to account for their feedback.
For example, thermal-hydraulics feedback plays an essential role in reactor operation.
Changes in the fuel temperature alter the cross sections due to Doppler broadening \cite{stacey2018nuclear}, which in turn impacts the neutron population and hence the fission rate and the rate of heat generation.
Multi-physics modeling accounts for these different feedback mechanisms; however, it poses a challenge in terms of the computational cost,
since it entails solving coupled systems for different fields such as neutron flux, temperature, and flow rate.
Moreover, the neutron transport (or diffusion) equation is no longer a linear one due to the indirect dependence of the cross section on the neutron flux.
several approaches are used to solve such systems.
The most commonly used one is a serial coupling of two or more separate codes in such a way that the output of one at a given time step, such as the flux distribution, is used as an input to the other, e.g., the temperature (which can be used then to update the cross section and so on).
This procedure is commonly referred to as operator splitting or explicit coupling.
One drawback of operator splitting is that time integration requires small steps to avoid stability issues, thus leading to high computational time. 
Alternatively, tight coupling in time between the physics can be maintained by using an implicit scheme that solves a non-linear iteration in each time step, also increasing the computational time.
A modern algorithm such as Jacobian free Newton-Krylov (JFNK) \cite{gaston2009parallel} can be used to perform this implicit coupling.
Another popular approach is to use fixed-point iteration, with an implicit time scheme to iterate between difference physics within a time step until convergence is satisfied, which is computationally taxing.
The aim of this work is to use a reduced-order model to overcome the high computational expense of these non-linear models.
In particular, the ROM was developed using the POD-Galerkin projection.
The method belongs to the projection-based family and can be described as a physics-driven method since it requires knowledge of the original governing equation as will be shown in Sec.\ref{sec:POD}.
In a previous work \cite{elzohery2021exploring}, we applied the method to the time-dependent diffusion equation where there was no coupling with other models and hence the application was linear.
In this study, the developed ROM was extended to treat the non-linearity induced by multi-physics modeling.
At this point, we must emphysize that the method used here, i.e; POD Galerkin projection has been used in previous works to approximate the diffusion equation \cite{} for different problems and codes where it showed promising results.
Moreover, the  Discrete Empirical Interpolation Method, DEIM, has been used in conjunction with the POD-Galerkin projection to approximate non-linear fields such as the temperature and the flow fields  .... \cite{}.  
Here, we are presenting another variant of the DEIM algorithm, known as the matrix version of DEIM, i,e; MDEIM, in order to improve the computational efficiency of the developed ROM.
The primary purpose of using the MDEIM is to obtain an offline-online decomposition of the ROM to avoid repetitive projection of the problem operator at each iteration within the time step.
The offline stage occurs only once and comprises the expensive step of the ROM.
On the other hand, the online stage involves solving a reduced number of equations at each time step with a computational cost independent of the FOM dimension and hence cheaper.
The LRA benchmark was used as a demonstrative application where the diffusion equation is coupled with a simplified thermal feedback model.

@@

%-----------------------------------------------------------------------------------------------------%
\section{The time-dependent diffusion equation}
\label{sec:fom}
In this work, the time dependent diffusion equation was used as the mathematical model of neutronic transients in nuclear reactors, i.e FOM, and for which an approximate ROM is sought.
The multi-group version of the equation is represented as
\begin{equation}
\begin{split}
\frac{1}{v_g} \frac{\partial\phi_g}{\partial t} &= 
\nabla. D_g\nabla\phi_g - \Sigma_{tg}\phi_g + \sum_{g^\prime=1}^{G}{\Sigma_{sg^\prime g}}\phi_{g^\prime} \\
&+ \chi_g(1-\beta)\sum_{g^\prime = 1}^{G}\nu\Sigma_{fg^\prime}\phi_{g^\prime}  
+ \chi_g\sum_{i=1}^{I} \lambda_i C_i \, ,
\end{split}
\label{time_diff}
\end{equation}
where 
\begin{equation*}
\frac{\partial C_i}{\partial t} = -\lambda_i C_i + \beta_i \sum_{g^\prime=1}^{G} \nu\Sigma_{fg^\prime} \phi_{g^\prime} \, ,
\end{equation*}
and $\phi_g$, $D_g$, $\Sigma_{ag}$, $\nu\Sigma_{fg}$ and $v_g$ are the neutron flux, diffusion coefficient, the absorption cross section, the production cross section and the average neutron velocity in group g, respectively. $\Sigma_{sg^\prime g}$ is the scattering cross section from group $g^\prime$ to group $g$ 
and $\chi_g$ is the probability that a fission neutron will be born in group $g$.
$C_i$ is the $i$th precursor concentration, $\beta_i$ is the fraction of delayed neutrons in group $i$, $\lambda_i$ is the $i$th group decay constant.
Upon spatial discretization, the resulting system of ODEs can be compactly represented in a matrix form as  
\begin{equation}
\frac{d\mathbf{y}(t)}{dt} = \mathbf{A}\mathbf{y}(t) \, ,
\label{eq:transient fom}
\end{equation}
where $\mathbf{y}\in\mathbb{R}^{N}$ is stacked vector of the flux and the precursors concentration, $\mathbf{A}\in\mathbb{R}^{N\times N}$ and $N = (I + G)\times n$ with $n$ is the number of spatial unknowns arising from discretization, $G$ is the number of energy groups, and $I$ is the number of precursors groups.
The operator $\mathbf{A}$ can be written in a block matrix form as
\begin{equation}
\mathbf{A} =  \left(\begin{array}{@{}c|c@{}}
\mathbf{L} & \mathbf{D} \\
\hline
\mathbf{F} & \mathbf{P}
\end{array}\right) \, .
\end{equation} 
where $\mathbf{L} \in \mathbb{R}^{(n\times G)\times (n\times G)}$ represents the diffusion operator,
$\mathbf{D} \in \mathbb{R}^{(n\times G)\times (n\times I)}$ is a block diagonal matrix representing the delayed neutrons production operator,
$\mathbf{P} \in \mathbb{R}^{(n\times I)\times ((n\times G)}$ is a diagonal matrix represents the precursors concentration loss operator, 
and $\mathbf{F}\in \mathbb{R}^{(n\times I)\times ((n\times I)}$ represents the precursors production operator.
%Thus, the system in Eq.~\ref{eq:transient fom} is the full-order model and it can be solved using a suitable time-scheme.
%However, it is expensive to solve such systems for fine spatial and time grids, since a linear system of a size proportional to the spatial discretization has to be solved for each time step.
%Hence, a reduced-order model to approximate it is justified.

@@

$\mathbf{\Phi}_1(t) = \mathbf{U}_1\mathbf{a}_1(t)$, $\mathbf{\Phi}_2(t) = \mathbf{U}_2\mathbf{a}_2(t)$ and $\mathbf{C}(t) = \mathbf{U}_c\mathbf{a}_c(t)$, where $\mathbf{U}_1$, $\mathbf{U}_2$  and $\mathbf{U}_c$ are the POD bases of the fast group, thermal group and precursor concentration respectively, and $\mathbf{a}_1$, $\mathbf{a}_2$ and $\mathbf{a}_c$ are their respective temporal coefficients. 
